---
title: "Using tidyverse for trade analysis"
subtitle: "School of Economics, Quaid-i-Azam University, Islamabad"
date: "`r format(Sys.time(), '%d %B %Y')`"
format: 
  html:
    toc: true
    #   code-fold: trueibliography: references.bib
---

## Trade Data Analysis

This material is mainly based on [UN ESCAPE online training on R for trade analysis](https://r.tiid.org/). In this post, `tidyverse` is used instead of base-R (used in the original material) due to convenience and reading codes like english language. Moreover,`ggplot2` provides awesome viusalization tools.

## Loading data

Data are provided on <https://r.tiid.org/> to be downloaded in `csv` file and one can read this data directly in R. I have save this data in R as `RData` and have loaded as follows. For more details on how to read various data formats in R, one can browse on how to import data. An easy way is File-\>Import Data Set -\> then choose relevant file. In case of error, upload relevant packages.

```{r}
#| warning: false
#| message: false
library(tidyverse)
load("trade_data.RData")

```

To get idea about the nature of this data, one may get `glimpse` ,`names` or use`head` etc.

```{r}
names(trade_data)
trade_data$value <- as.numeric(as.character(trade_data$value)) # trade value as numeric
# drop missing values
trade_data <- trade_data[!is.na(trade_data$value),]
trade_data
```

We notice that there are **5 main variables**, namely: **reporter** indicating destination country, **flow** consists of import and export , **year** for which trade data are included in this file and **value** is USD.\`

We can now check the **`value`** in billion by dividing the previous function by a billion. Notice after the comma we specify the value column. Other than very large exporting/importing countries, one may convert **`value`** in millions of USD instead of taking it in billions of USD if required.

```{r}
trade_data %>% distinct(year)  # distinct number of years for which this trade data are available
trade_data<-trade_data %>% mutate(value=value/1000000000)

options(scipen = 999) ## To avoid scientific numbers (like appearing in exponents)
```

## Pakistan

Now we analyze data for Pakistan for year **2016** and learn R for data exploration in this post.

```{r}

trade_data %>% filter(flow=="E"&reporter==564&year==2016&partner==1) %>% select(value)

```

So total export by Pakistan to the world (world coder in partner category is 1) is USD 20.8 billion. Now if we calculate Pakistan's export with all countries, it is `r trade_data %>% filter(flow=="E"&reporter==564&year==2016&partner!=1) %>% summarise(sum(value))` USD Billion . This is greater than we calculated earlier with world. This is because there are many categories in partner category which include countries as well as different regions. So need some data wrangling before one can match both the figures. One has to exclude all those partner codes which are not equal to 1 and **isCountry==1**.

## Export plot for all the years

```{r}
library(lubridate)
PK_X<-trade_data %>% filter(flow=="E"&reporter==564&partner==1)
PK_X$year<-as_date(PK_X$year)
ggplot(PK_X)+aes(x=year,y=value)+geom_line()+labs(y="USD, billions", title = "Pakistan's world export")
```

## Import and Export Together

```{r}
PK_XM<-trade_data %>% filter(reporter==564&partner==1)
PK_XM$year<-as_date(PK_XM$year)
ggplot(PK_XM, aes(x = year, y = value)) + 
  geom_line(aes(color = flow, linetype=flow)) + 
  scale_color_manual(values = c("darkred", "steelblue"))
```

## Top 10 exporting destination for China

Let's now show in a plot the top 10 countries Pakistan exported to in 2016.

We start by creating a table with Pakistanis exports in 2016 to individual countries:

```{r}
PK_X_countries <- trade_data %>% filter(flow=="E"&reporter==564&year==2016&partner==1)

```

But to get top trade partners we have to extract names of the countries for which following file is downloaded from the IMF. **Partner** codes are available but country names are not given which one has to download from the IMF website.

## Country code

To select countries, \`isCountry" or code has to be set 1 so other categories like advanced countries and countries not specified are excluded

```{r}

imf_codes <- read_csv("imf_codes.csv")
codes<-imf_codes %>%  select(code,country,isCountry, region)
cntry_w <- codes %>% filter(isCountry==1|code==1) %>% select(code)
cntry_w
```

```{r, eval=FALSE}
PK_X_2016<-trade_data %>% filter(flow=="E"&reporter==564&year==2016&partner!=1&partner %in% cntry_w) 
PK_X_2016 
PK_X_2016<-trade_data[trade_data$flow=="E"& #select only exports
trade_data$reporter==564& #from china
trade_data$year==2016& #in 2016
trade_data$partner!=1& #where country is not world (code for world is 1)
trade_data$partner %in% cntry_w, #which is in the list of codes that we filtered earlier
"value" #variable we're interested in
]
sum(PK_X_2016)/1000000000
```

## Merge country codes and trade_data

Lets do the same exercise for Pakistan before proceeding further.

```{r}
PK_X_countries <-trade_data %>% filter(flow=="E"&reporter==564&year==2016&partner!=1)

PKX_merged<-left_join(PK_X_countries,codes,by=c("partner"="code"))
PKX_merged<-PKX_merged %>% select(country, value, isCountry,region)
PKX_merged
# Only keep countries
PKX_merged<-PKX_merged %>% filter(isCountry==1)

```

## Top 10 exporting destinations in case of Pakistan in 2016

Sort export value by descending order using `arrange()` and then select top 10 exporting destinations.

```{r}
PKX_merged_top10<-PKX_merged %>% arrange(desc(value)) 
PKX_merged_top10<-head(PKX_merged_top10,10)
PAK<-ggplot(PKX_merged_top10, aes(x = reorder(country,-value), y = value*1000, fill=region))+geom_bar(stat='identity')+theme(axis.text.x = element_text(angle = 90, size = 10))+labs(x="",y="USD Millions",title="Top 10 exporting destinations of Pakistan in 2016")

PAK
```

## India

```{r, eval=FALSE}
PK_X_2016<-trade_data %>% filter(flow=="E"&reporter==564&year==2016&partner!=1&partner %in% cntry_w) 
PK_X_2016 
PK_X_2016<-trade_data[trade_data$flow=="E"& #select only exports
trade_data$reporter==564& #from china
trade_data$year==2016& #in 2016
trade_data$partner!=1& #where country is not world (code for world is 1)
trade_data$partner %in% cntry_w, #which is in the list of codes that we filtered earlier
"value" #variable we're interested in
]
sum(PK_X_2016)/1000000000
```

## India

```{r}
#| echo: false
IN_X_countries <-trade_data %>% filter(flow=="E"&reporter==534&year==2016&partner!=1)

INX_merged<-left_join(IN_X_countries,codes,by=c("partner"="code"))
INX_merged<-INX_merged %>% select(country, value, isCountry,region)
INX_merged<-INX_merged %>% filter(isCountry==1)
INX_merged_top10<-INX_merged %>% arrange(desc(value)) 
INX_merged_top10<-head(INX_merged_top10,10)
IND<-ggplot(INX_merged_top10, aes(x = reorder(country,-value), y = value*1000, fill=region))+geom_bar(stat='identity')+theme(axis.text.x = element_text(angle = 90, size = 10))+labs(x="",y="USD Millions",title="Top 10 exporting destinations of India in 2016")
IND
```

##Bangladesh

```{r}
#| echo: false
BD_X_countries <-trade_data %>% filter(flow=="E"&reporter==513&year==2016&partner!=1)

BDX_merged<-left_join(BD_X_countries,codes,by=c("partner"="code"))
BDX_merged<-BDX_merged %>% select(country, value, isCountry,region)
BDX_merged<-BDX_merged %>% filter(isCountry==1)
BDX_merged_top10<-BDX_merged %>% arrange(desc(value)) 
BDX_merged_top10<-head(BDX_merged_top10,10)
#BDX_merged_top10
BD<-ggplot(BDX_merged_top10, aes(x = reorder(country,-value), y = value*1000, fill=region))+geom_bar(stat='identity')+theme(axis.text.x = element_text(angle = 90, size = 10))+labs(x="",y="USD Millions",title="Top 10 exporting destinations of Bangladesh in 2016")
BD

```

## Sri Lank

```{r}
#| echo: false
SL_X_countries <-trade_data %>% filter(flow=="E"&reporter==524&year==2016&partner!=1)

SLX_merged<-left_join(SL_X_countries,codes,by=c("partner"="code"))
SLX_merged<-SLX_merged %>% select(country, value, isCountry,region)
SLX_merged<-SLX_merged %>% filter(isCountry==1)
SLX_merged_top10<-SLX_merged %>% arrange(desc(value)) 
SLX_merged_top10<-head(SLX_merged_top10,10)
SRL<-ggplot(SLX_merged_top10, aes(x = reorder(country,-value), y = value*1000, fill=region))+geom_bar(stat='identity')+theme(axis.text.x = element_text(angle = 90, size = 10))+labs(x="",y="USD Millions",title="Top 10 exporting destinations of Sri Lank in 2016")
SRL
```

## South Asia

Total export of Bangladesh, India, Pakistan and Sril Lanka in billion of USD is `r trade_data %>% filter(flow=="E"&reporter==513&year==2016&partner==1) %>% summarise(sum(value))`, `r trade_data %>% filter(flow=="E"&reporter==534&year==2016&partner==1) %>% summarise(sum(value))`, `r trade_data %>% filter(flow=="E"&reporter==564&year==2016&partner==1) %>% summarise(sum(value))` and `r trade_data %>% filter(flow=="E"&reporter==524&year==2016&partner==1) %>% summarise(sum(value))` respectively.

```{r}
library(gridExtra)
grid.arrange(BD,IND,PAK,SRL, ncol=2)
```
